Inferring the gravitational potential of the Milky Way with a few precisely measured stars

### 2014-03-11 text/paper2/ms.tex
minor notes

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, and evolved to present day in the parent potential, $\Phi$. That is, we assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} We add two other parameters: a constant offset, $\Loffset$, that specifies the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail.
\begin{align}
	p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit) &= p(\bs{r} \given \bs{r}_\sat, \tub, \Phi, \Loffset, \tailbit)\,p(\bs{v} \given \bs{v}_\sat, \tub, \Phi)\,\left\vert\J\right\vert_{t=\tub}\\
	p(\bs{r} \given \bs{r}_\sat, \tub, \Phi, \Loffset, \tailbit) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)_{t=\tub}\\
	p(\bs{v} \given \bs{v}_\sat, \tub, \Phi) &= \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. This model could trivially allow for mass loss by incorporating the initial satellite mass, $m$ and, e.g., a constant mass-loss term, $\dot{m}$. For a single star, in detail, the model works as follows:
\begin{enumerate}
	\item sample from the observed distributions in heliocentric coordinates (Eq.~\ref{eq:obsstar} and Eq.~\ref{eq:obsprog}) to get present-day positions for both the star and progenitor;
	\item transform to Galactocentric, cartesian coordinates, and integrate both orbits backwards (as test particles) to $t=t_{\rm ub}$;
	\item evaluate the likelihood.
\end{enumerate}

This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking (cite?). We will explore properly modeling this process in future work. For this article, we modify the above to be a mixture model: we add a new parameter for each star, $P_{\rm shock}$, which is the probability that the star is drawn from a broader, ``shocked'' distribution instead of a tight Gaussian centered on an effective Lagrange point:
\begin{multline}
	p_{\rm strip} = [ \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)\\
		\times \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v) \left\vert\J\right\vert ]_{t=\tub}
\end{multline}
\begin{align}
	p_{\rm shock} = \left [ \mathcal{N}(\bs{r} \given \bs{r}_\sat, 9\sigma^2_r)\mathcal{N}(\bs{v} \given \bs{v}_\sat, 9\sigma^2_v) \left\vert\J\right\vert\right]_{t=\tub}
\end{align}
\begin{multline}
	p(\W \given \W_\sat, \tub, \Phi, \tailbit, c, P_{\rm shock}) = \\
		P_{\rm shock}p_{\rm shock} + (1-P_{\rm shock})p_{\rm strip}
\end{multline}
The full posterior probability for the position of a star, progenitor, and properties of the parent potential, given the data, is then
\begin{multline}
	p(\Phi, \W, \W_\sat, \tub, \tailbit, c, P_{\rm shock} \given \D, \D_\sat) = \\
		\shoveleft{\hspace{3em}\frac{1}{\mathcal{Z}} \, p(\D \given \W) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit) \, 
			p(\Phi) \, p(\tub) \, p(c)
\end{multline}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. All model parameters are summarized in Table~\ref{tbl:params}.

@@

>>>>

To model the true, 6D, present-day position of the star, $\W$, we first transform to cartesian, Galactocentric coordinates where ($\bs{r},\bs{v}$) and ($\bs{r}_\sat,\bs{v}_\sat$) are the position and velocity of the star and progenitor. In cartesian coordinates, the present-day position of the star is taken as having been sampled at $t=\tub$ from an isotropic Gaussian centered on one of the instantaneous, effective Lagrange points in position, and an isotropic Gaussian centered on the origin in velocity, and evolved to present day in the parent potential, $\Phi$. That is, we assume that once the star becomes ``unbound'' from the satellite the potential of the satellite can be ignored, and thus treat the star as a test particle.\footnote{This assumption breaks down for sufficiently large-mass progenitors, somewhere between $\sim10^8-10^9~$M$_\odot$.} We add two other parameters: a constant offset, $\Loffset$, that specifies the position of the effective Lagrange points relative to the classical tidal radius, and a binary parameter $\tailbit$, that is either $-1$ or $+1$ depending on whether the star is in the leading or trailing tail.
\begin{align}
	p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit) &= p(\bs{r} \given \bs{r}_\sat, \tub, \Phi, \Loffset, \tailbit)\,p(\bs{v} \given \bs{v}_\sat, \tub, \Phi)\,\left\vert\J\right\vert_{t=\tub}\\
	p(\bs{r} \given \bs{r}_\sat, \tub, \Phi, \Loffset, \tailbit) &= \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)_{t=\tub}\\
	p(\bs{v} \given \bs{v}_\sat, \tub, \Phi) &= \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v)_{t=\tub}
\end{align}
where $\rtide$ is given by Eq.~\ref{eq:tidalradius}, and $\sigma_v$ by Eq.~\ref{eq:velscale}. $\left\vert\J\right\vert$ is the absolute value of the determinant of the Jacobian that defines the transformation from heliocentric, spherical to Galactocentric, cartesian coordinates. This model could trivially allow for mass loss by incorporating the initial satellite mass, $m$ and, e.g., a constant mass-loss term, $\dot{m}$. For a single star, in detail, the model works as follows:
\begin{enumerate}
	\item sample from the observed distributions in heliocentric coordinates (Eq.~\ref{eq:obsstar} and Eq.~\ref{eq:obsprog}) to get present-day positions for both the star and progenitor;
	\item transform to Galactocentric, cartesian coordinates, and integrate both orbits backwards (as test particles) for the total interaction time, $\tint$;
	\item evaluate the likelihood at $t=\tub$.
\end{enumerate}

This model works well when all stars are lost due to tidal pruning. Even on a mildly eccentric orbit, stars become preferentially unbound around the satellite's pericenter because of tidal shocking (cite?). We will explore properly modeling this process in future work. For this article, we modify the above to be a mixture model: we add a new parameter for each star, $P_{\rm shock}$, which is the probability that the star is drawn from a broader, ``shocked'' distribution instead of a tight Gaussian centered on an effective Lagrange point:
\begin{multline}
	p_{\rm strip} = [ \mathcal{N}(\bs{r} \given \bs{r}_\sat - \Loffset\,\tailbit\,\rtide\,\hat{x}_1, \sigma^2_r)\\
		\times \mathcal{N}(\bs{v} \given \bs{v}_\sat, \sigma^2_v) \left\vert\J\right\vert ]_{t=\tub}
\end{multline}
\begin{align}
	p_{\rm shock} = \left [ \mathcal{N}(\bs{r} \given \bs{r}_\sat, 10\sigma^2_r)\mathcal{N}(\bs{v} \given \bs{v}_\sat, 10\sigma^2_v) \left\vert\J\right\vert\right]_{t=\tub}
\end{align}
\begin{multline}
	p(\W \given \W_\sat, \tub, \Phi, \tailbit, c, P_{\rm shock}) = \\
		P_{\rm shock}p_{\rm shock} + (1-P_{\rm shock})p_{\rm strip}.
\end{multline}
Here we assume that the ``shocked'' distribution has a much larger variance, and arbitrarily set this to 10 times the variance of the stripped debris. The full posterior probability for the position of a star, progenitor, and properties of the parent potential, given the data, is then
\begin{multline}
	p(\Phi, \W, \W_\sat, \tub, \tailbit, c, P_{\rm shock} \given \D, \D_\sat) = \\
		\shoveleft{\hspace{3em}\frac{1}{\mathcal{Z}} \, p(\D \given \W) \, p(\D_\sat \given \W_\sat)}\\
			\times p(\W \given \W_\sat, \tub, \Phi, \Loffset, \tailbit) \, 
			p(\Phi) \, p(\tub) \, p(c)
\end{multline}
where the factor $\mathcal{Z}$ only depends on the properties of the data (the \emph{evidence}), and is thus constant when varying the model parameters. All model parameters are summarized in Table~\ref{tbl:params}.

@@

We first
